---
title: "edfReader vignette"
author: "Jan Vis"
date: "3-3-2018"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{edfReader vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
# 1. Introduction
## 1.1 EDF and BDF
The [European Data Format](http://www.edfplus.info) (EDF) is a simple and flexible format for exchange and storage of multichannel biological and physical signals.  It was developed by a few European 'medical' engineers who first met at the 1987 international Sleep Congress in Copenhagen. See http://www.edfplus.info/

The original EDF specification has been expanded in several ways. EDF+ supports the addition of annotations and non-continuous recordings. The [BioSemi Data Format](http://www.biosemi.com/faq/file_format.htm) BDF format uses 24 bits per sample (in stead of the 16 bits per sample in EDF). And [BDF+](http://www.teuniz.net/edfbrowser/bdfplus%20format%20description.html) is an EDF+ like extension of BDF.

This packages supports full reading of all these variants.

Both EDF and BDF files consist of an header followed by the data records with one or more signals recorded, either ordinary signals or annotation signals.

This package follows this structure by providing two basic functions: readEdfHeader and readEdfSignals (see the help pages for details)

## 1.2 Sample files
The examples below are based on the following files: 

- AFile an EDF+ file with a continuous recording of 2 ordinary signals, two annotation signals and a sub second start specification

- BFile a BDF+ file with a continuous recording of 11 ordinary signals and one annotation signal   

- CFile an EDF+ file with a continuous recording of 11 ordinary signals and one annotation signal 

- DFile an EDF+ file with a discontinuous recording of 11 ordinary signals and one annotation signal 

The BFile, CFile and DFile are derived from the "test_generator_2" test files from http://www.teuniz.net/edf_bdf_testfiles.
The AFile is derived from the "test_generator8" file received from Teunis van Beelen by private communications.

```{r}
libDir <- system.file ("extdata", package="edfReader")
AFile <- paste (libDir, '/edfAnnonC.edf', sep='') # a file with 2 annotation signals
BFile <- paste (libDir, '/bdfPlusC.bdf' , sep='') # a continuously recorded BDF file
CFile <- paste (libDir, '/edfPlusC.edf' , sep='') # a continuously recorded EDF file
DFile <- paste (libDir, '/edfPlusD.edf' , sep='') # a discontinuously recorded EDF file
```

# 2. EDF headers objects
## 2.1 Introduction
The readEdfHeader function returns a list of class 'ebdfHeader' with all the data from the EDF or BDF file header.
Part of this list is a data frame of class 'ebdfSHeader' which contains the signal headers.

In this chapter and the next, the EDF data read by edfReader is explained by using the S3 print() and summary() functions. The summary() provides somewhat more information than the print() function. In general, print() doesn't contain internal EDF/BDF file data (like, e.g. the EDF signal number). This type of data is displayed with a summary() only.

Both print() and summary() may be used with a maxRows and/or file parameter. 'maxRows' will limit the number of rows in tables, its defult value = 24. 'file' will redirect the output to the file specified.

## 2.2 The ebdfHeader
A file header can be read with readEdfHeader()
```{r}
require (edfReader)
AHdr  <- readEdfHeader (AFile)
BHdr  <- readEdfHeader (BFile)
CHdr  <- readEdfHeader (CFile)            
DHdr  <- readEdfHeader (DFile)                  
```

Summaries of the header data can be shown with the S3 summary () and print() functions
```{r}
BHdr
summary (AHdr)
```
edfAnnonC.edf contains two annotation. These signals must have the same label. As edfReader names signals after their labels (see below), distinguished names are created by appending '-1' and '-2' to the labels.

NOTE \ \ \ \ \ \ Actually, the startTime fraction .7 is not read from the EDF+/BDF+ file header but from the first data record. For details, see the section “Samples, time and periods”.

## 2.3 The ebdfSHeader
The print() and summary() output for an ebdfSHeader is shown below.
```{r}
AHdr$sHeader
summary (CHdr$sHeader)
```

# 3. EDF signal objects
## 3.1 Introduction
The signals in an EDF or BDF file can be read with the readEdfSignals function.

The readEdfSignals function with simplify=FALSE returns a list of class 'ebdfSignals' with the signals selected from the EDF / BDF file. 

The signals in this list are of the following :

- continuous signals   
of class 'ebdfCSignal' which may be from all types of EDF(+C)/BDF(+C) files

- fragmented signals      
of class 'ebdfFSignal' which may be from EDF+D and BDF+D files only

- annotation signals    
of class 'ebdfASignal’ which may be from all types of EDF+ and BDF+ files.

## 3.2 Reading the whole recording of all signals
### 3.2.1 Reading a file with continuously recorded signals
```{r}
ASignals <- readEdfSignals (AHdr)
summary (ASignals)
```
NOTE 1\ \ \ \ \ In case there is only one ordinary signal or annotation signal the data is displays in a number lines (see the annotation signal), if not, a table is used (see the ordinary signals).

NOTE 2\ \ \ \ \ The EDF signal number '1, 3' indicates that the annotation object contains the annotations from both the EDF signals 1 and 3. This can be prevented by using the 'mergeASignals=FALSE' option.

NOTE 3\ \ \ \ \ Because of the merging of the annotation signals, the R signal numbers differ from the EDF signal numbers.

NOTE 4\ \ \ \ \ 'Last end' equals the maximum of the annotations onsets + durations (see the section 'annotation data' below).

### 3.2.2 Reading a file with discontinuously recorded signals as a single signal
```{r}
DSignals <- readEdfSignals (DHdr)
DSignals
```
NOTE 1\ \ \ \ \ By default all fragments in a discontinuously recorded signals will be concatenated into one signal with the gaps filled with NA values. This can be prevented by using the 'fragments=TRUE' option.

NOTE 2\ \ \ \ \ The number of NAs equals the signal length minus the number of samples.

NOTE 3\ \ \ \ \ As the duration was not specified in the EDF file, the value of 'Last end' equals NA (see the section 'annotation data' below).

### 3.2.3 Reading a file with discontinuously recorded signals as a list of fragments
```{r}
DSignalsF <- readEdfSignals (DHdr, fragments = TRUE)
summary (DSignalsF)
```

NOTE \ \ \ \ \ \ As a fragment doesn't contain NAs there is no need for a 'signal length' column in the table. In stead the number of fragments per signal is provided.

## 3.3 Reading a selection of signals

The reading of signals can be restricted to a specific set by using the signals parameter. Signals can be identified by their EDF signal number, label, name, or signal type ('Ordinary' or 'Annotations'). Also, the list may contain duplicates. The following 3 designations refer e.g. to the same signal.

```{r}
CSignals8 <- readEdfSignals (CHdr, signals=c(8, "8", "sine 8.5 Hz"))
summary (CSignals8)
```

NOTE \ \ \ \ \ \ As in this case only one signal has been read, the list of one signal was simplified to a single 'ebdfCSignal' object with R signal number 1. This could be prevented by using the 'simplify=FALSE' parameter.

EXAMPLE \ \ \ \ \ \ readEdfSignals (CHdr, signals=7, simplify=FALSE)[[1]] and readEdfSignals (CHdr, signals=7) will return the same object.

## 3.4 Reading a selected period
If required the reading can be restricted to a particular period. A period is identified with the 'from' and 'till' parameters which specify a time in seconds relative to the start of the recording.
```{r}
ASignalsPeriod    <- readEdfSignals (AHdr, from=0.7, till=1.8)
summary (ASignalsPeriod)
```
NOTE 1 \ \ \ \ Because the recorded period is only 1.2 sec (see above or use summary(ASignalsPeriod)), the period read for ordinary signals will not be  [0.7, 1.8) but only [0.7, 1.2).

NOTE 2 \ \ \ \ For time and rounding details see the section “Samples, time and periods”

NOTE 3 \ \ \ \ The annotations included are those for which the period [onset, onset+duration] overlaps with the [from, till] period  (see the section 'annotation data' below).

## 3.5 Ordinary signals, continuously recorded
```{r}
CSignals <- readEdfSignals (CHdr)
summary (CSignals$pulse)         # edfReader names signals after their label
```
## 3.6 Ordinary signals, not continuously recorded
Ordinary signals that are not continuously recorded can be read in two different ways:

- as a 'continuous' sequence of samples with 'NA' values filling the gaps (use 'fragments=FALSE', the default).       
The result is of class 'ebdfCSignal' (the same as for a continuously recorded signal).

- as a number fragments of continuously recorded parts (use 'fragments=TRUE')      
The result is of class 'ebdfFSignal'.

The latter method will use a more complex data structure, the first may result in a (much too) huge object.
```{r}
CDSignals <- readEdfSignals (DHdr, from=5.1, till=18)
FDSignals <- readEdfSignals (DHdr, fragments=TRUE)
```
The objects of class 'ebdfCSignal' are printed and summarised in the same way as continuously recorded signals.
```{r}
summary (FDSignals$`sine 8.5 Hz`)         # note the "`" quotes for a name with spaces.
```

## 3.7 Annotation signals
```{r}
CSignals$`EDF Annotations`
```
NOTE \ \ \ \ \ \ "= ... h:m:s" is added to a time value if that value is greater than 300 seconds. 
```{r}`
summary (ASignalsPeriod$`EDF Annotations`)
```
NOTE 1 \ \ \ \ Because both annotation signals are merged into one, the merged signal is named `EDF Annotations` again.

NOTE 2 \ \ \ \ The "Record start specs = 0" indicates that the record start specifications are not included, i.e. readEdfSignals was used with the parameter recordStartTimes = FALSE.


# 4. Samples, time and periods
## 4.1 The start time
In case of an EDF or BDF file, the startTime attribute is based on the startdate and starttime in the EDF/BDF header.

The starttime in an EDF/BDF header is specified up to the second.

In EDF+ and BDT+ files a sub second start time can be specified as the start time of the first data record. See the last paragraph in  section 2.2.4 of the EDF+ specification.

See, e.g., the startTime for the AFile:
```{r}
format (AHdr$startTime, format="%Y-%m-%d %H:%M:%OS3",  usetz = FALSE)
```
For EDF+/BDF+ files, the startTime shown by the edfReader is based on both the data in the header and the EDF+/BDF+ start time of the first data record, i.e. on the actual start of the recording. As a consequence the R start time of this first data record is always 0.
```{r}
ASignalsPlusStartTimes <- readEdfSignals(AHdr, signals='EDF Annotations-1', recordStarts=TRUE)
ASignalsPlusStartTimes$recordStartTimes$startTime[1]
```

By the same token, the onset for annotations is relative to this R startTime (in an EDF+/BDF+ file the onset is relative to the startdate and starttime in the EDF/BDF header).

## 4.2 Samples and time
As usual, a recording starts at time 0 (relative to the recording startTime) and with sample 1.

Consequently sample n will be at time (n-1)/sRate, where sRate denotes the sample rate.

## 4.3 Samples and periods
Apart from rounding errors, a from - till period in readEdfSignals will be the period [from,till), i.e. starting at from and up to be but not including till.

This may sound strange, but this convention has the following properties:

- the sample rate can be defined as the total number of samples divided by the total time

- or, more precise, any period which is a multiple of the sample period sPeriod will always contain the same number of samples     
i.e. for any t ≥ 0, [t, t+n*sPeriod) will always contain n samples

- for any number of adjacent periods, the total number of samples will be always the sum of the samples in the individual periods.

- the first sample will be ceiling(from*sRate) + 1

- the last sample will be ceiling(till*sRate)

- with from=t and till=t, the empty set [t,t) corresponds to [ceiling(t\*sRate) + 1, ceiling(t\*sRate)] which is also empty.

## 4.4 Alignment
### 4.4.1 The issue
In an EDF+/BDF+ file the start of the recording of the signals in a data record is specified in its first annotations signal. For a +D file the recording in a subsequent 
data record may start at any time after the start of the previous data record plus the record duration time. 

However, the gap between to +D file data records may not be an exact multiple of the sample period of its signals. This raises a question about the alignment of samples in these subsequent data records. 

### 4.4.2 Two basic models.
The alignment of ordinary signals can be modelled in two different ways:

A.  with a interrupted clock     
In which case the first sample for every recorded signal is taken at the start of the recording for that data record

B.  which a continuously running clock      
In which case the clock starts at the start of the first record and all sampling is based on the individual signal sample rate(s). In other words,
the sample time for a signal sample in the data records is aligned to its sample rate.

It should be noted that the model with the continuously running clock is (implicitly) required if one want to map all fragments of a signal into a single single (with NA values in the gap) one.

The edfReader supports both models, for any fragment it provides both the aligned data according to the "continuously running clock" model as well as the record start time from the data record.

### 4.4.3 Alignment details
The first sample in every +D record can be aligned as follows:
    
    n = ceiling (sRate * recordStartTime) + 1

and the 

    sampleTime = (n - 1) / sRate
    
Where  

- sRate is the sample rate

- n is the sample number when continuously sampled from the start of the recording

- recordStartTime denotes the start of the record relative to the start of the recording

- sampleTime the (aligned) sample time for the n's sample

However, in order to avoid rounding unnecessary rounding errors the first formula is actually implemented as follows:

    n = ceiling (sRate * (recordStartTime - 2 * distanceToNextLowerNumeric)) + 1
    
Where
    distanceToNextLowerNumeric = the distance between recordStartTime and its next lower numeric (double float) value.
    
NOTE \ \ \ \ \ \ In earlier edfReader versions the margin was 5 * .Machine\$double.eps. And .Machine\$double.eps equals the distanceToNextLowerNumeric if x == 2. See e.g. https://en.wikipedia.org/wiki/Machine_epsilon for details.

## 4.5 Parameters involved

For any signal object and any fragment:

- the 'start' parameter contains the aligned signal / fragment start time in seconds, relative to the start in the recording (for a signal it is always 0)

- the 'fromSample' parameter contains the sample number for the corresponding sample

- the 'recordingStart' parameter contains the start of the recording for the data record. For a '+' file this is the time specified in the first annotation signal

- a 'from' parameter in an readEdfSignals invocation is relative to the start of the recording. The default value is 0

- a 'till' parameter in an readEdfSignals invocation is relative to the start of the recording. The default value is Inf

- a 'recordedPeriod' parameter equals the number of data records times the record duration time as specified in the header

- a totalPeriod parameter contains the period from the start time in the header till the end of the recording of the ordinary signals, i.e. the record start time of the last data record plus the record duration time

# 5. Object details
## 5.1 Introduction
In this chapter the strucnture of the R EDF/BDF objects is explained by using the str(), or more often, the str1() function as defined below.
```{r}
str1 <- function (x) str (x, max.level = 1)
```
NOTE \ \ \ \ \ \ The str() function may round numeric and POSIXct values !

## 5.2 Header details
### 5.2.1 Header attributes
The header data encompass the following:
```{r}
str1 (CHdr)
```
The fields version, patient, recordingId, startTime, headerLength, reserved, nRecords, recordDuration, and nSignals are from the file header. 

the startSecondFraction attribute contains the sub second start data specified in the first data record. See the 'The start time' section above.

sHeaders is detailed below. The others attributes are derived ones.

### 5.2.2 Signal header attributes
The signal header data encompass the following:
```{r}
str1 (CHdr$sHeader)
```
The fields label, transducerType, physicalDim, physicalMin, physicalMax, digitalMin, digitalMax, preFilter, samplesPerRecord, and reserved are from the file header. The others are derived. Gain and offset are used to map the digital sample values to the range of physical values.

For annotation signals the only relevant fields are "label" which must have the value "EDF Annotations" (or "BDF Annotations") and "samplesPerRecord".

## 5.3 Signal details
### 5.3.1 Ordinary signal objects of class 'ebdfCSignal'
The data for ordinary signal objects of class 'ebdfCSignal' encompass the following:
```{r}
str1 (CSignals$pulse) 
```
The attributes startTime, signalNumber, label, isContinuous, isAnnotation, recordedPeriod, totalPeriod, transducerType, sampleBits, sRate, range, and preFilter are (derived) from the header data.

The signalNumber denotes the EDF signal number, i.e. the signal number in the EDF/BDF file. The RSignalNumber denotes the number / index in the R list of signals read.

For a continuously recorded signal the totalPeriod is equal to the recordedPeriod. For a not continuously recorded signal the total period equals the start of the last data record plus its duration.

The attributes from and till contain the values of the corresponding actual readEdfSignals parameters. The default values are 0 and Inf.

The start attribute contains the start time and is always zero. The fromSample attribute contains the first sample number and is always 1. By including these attributes, signals and fragments share the same time/sample attributes.

The signal attribute contains the sample data from the EDF / BDF data records. If read with the readEdfSignals parameter physical=TRUE, the default, the digital sample values are mapped to physical values. With physical=FALSE, signals contains the digital sample values.

The physical values are calculated as follows:   
    physicalValue = gain * digitalValue + offset,    
    with:
    gain   = (physicalMax - physicalMin) / (digitalMax  - digitalMin)     
    offset = physicalMax - gain * digitalMax     

### 5.3.2 Ordinary signal objects of class 'ebdfFSignal'
#### 5.3.2.1 The fragmented signal
The data for ordinary signal objects of class 'ebdfFSignal' encompass the following:
```{r}
str1 (FDSignals$`sine 8.5 Hz`) 
```
For the attributes startTime, signalNumber, label, isContinuous, isAnnotation, recordedPeriod, totalPeriod, from, till, start, fromSample, transducerType, sampleBits, sRate, range and preFilter see the previous section.

For a not continuously recorded signal the total period equals the start of the last data record plus its duration.

The fragments attribute contains the list of recorded fragments.

### 5.3.2.2  The fragments
The data of a signal fragment in objects of class 'ebdfFSignal' encompass the following:
```{r}
str1 (FDSignals$`sine 8.5 Hz`$fragments[[1]]) 
```
The fromSample attribute contains the sample number of the first sample in this fragment (as if it were a continuous recording). The start attribute the sample time for this sample.

The recordingStart attribute contains the start of the recording as specified in the data racored (minus the start time of the first data record, see the section 'The start time').

The 'signal' attribute contains the fragment's sample values. These may be physical values (the default) or digital values (see above).

### 5.3.3 Annotation signals
#### 5.3.3.1 The annoration signal
The data for annotation signal objects of class 'ebdfASignal' encompass the following:
```{r}
ASignals2 <- readEdfSignals (AHdr, recordStarts=TRUE)
str1 (ASignals2$`EDF Annotations`) 
```
For the attributes startTime, signalNumber, label, isContinuous, isAnnotation, totalPeriod, from, and till see the section for objects of class 'ebdfCSignal'.

The annotations attribute contains a data frame with the individual annotations.

NOTE \ \ \ \ \ \ The POSIXct startTime has been rounded by str().

### 5.3.3.2 The annotations
The data for a single annotation encompass the following:
```{r}
str1 (ASignals2$`EDF Annotations`$annotations) 
```
The record attribute refers to the data record the annotation was read from.

The onset attribute contains the time of the annotation relative to the start of the recording.

The duration attribute contains the duration of the annotated event. The duration may not be specified in the EDF/BDF file.

The annotation attribute contains the annotation associated with the onset and duration.

The end attribute equals the sum of the onset plus the duration.

The fromSignal attribute, if present, refers to the signal that contains the annotation. This attributed is present only if the annotations from different signals were merged into one ‘ebdfASignal’ object. 

NOTE \ \ \ \ \ \ In case readEdfSignals is used with a from and till parameter only those annotatins are included for which the onset is within the [from, till] interval or - in case the duration is specified - the periodd [onset, onset+duration] overlaps with this interval.

### 5.3.3.3 The record start times
In an EDF+/BDF+ file the start of the recording of a data recorded is specified in the (first) annotation signal. When read with recordStarts=TRUE, these start times are included in the recordStartTimes data frame.

```{r}
str1 (ASignals2$`EDF Annotations`$recordStartTimes) 
```
NOTE \ \ \ \ \ \ The startTime is relative to the R startTime, i.e. the start of the recording. (in an EDF+/BDF+ file the onset is relative to the startdate and starttime in the EDF+/BDF+ header).

# 6. Next step: a  quick look
One of the the first things you may want to do with the imported signals is to have look at them.

This can be achieved with one of the several plot packages available, e.g.:

	- ‘plot’ (included in the base installation)

	- ‘lattice', or

	- 'ggplot2' 

As an example, the function plotEdfSignals() uses ggplot2 and will plot one or more signals  
over some period of time. The plot shown presents the period from 0.2 till 0.5 seconds of the 
'sine 8.5 Hz' and the 'sine 50 Hz' signals in CSignals.

```{r, fig.width=7.2, fig.height=2.5}
plotEdfSignals <- function (signals,labels, from=0, till=Inf) {
    nLabels <- length (labels)
    sRate   <- numeric (length = nLabels)
    fromS   <- integer (length = nLabels)
    tillS   <- integer (length = nLabels)
    sLength <- integer (length = nLabels)
    for (i in 1:nLabels) {
        sRate[i]    <- signals[[labels[i]]]$sRate
        fromS[i]    <- ceiling (sRate[i] * max (from, 0)) +1
        tillS[i]    <- ceiling (sRate[i] * till)
        tillS[i]    <- min (tillS[i], length(signals[[labels[i]]]$signal))
        sLength[i]  <- tillS[i] - fromS[i] + 1 
    }
    totLength  <- sum (sLength)
    cat (" totLength=",  totLength)
    time    <- numeric   (length = totLength)
    signal  <- numeric   (length = totLength)
    label   <- character (length = totLength)
    from <- 1
    for (i in 1:nLabels) {
        till <- from + sLength[i] - 1
        time  [from:till]   <- seq (from=fromS[i]-1, to=(tillS[i]-1)) / sRate[i]
        signal[from:till]   <- signals[[labels[i]]]$signal[fromS[i]:tillS[i]]
        label [from:till]   <- rep(labels[i], sLength[i])
        from <- till + 1
    }
    cat (" | from-1=", from-1,'\n')
    
    ggplotDF <- data.frame (time=time, signal=signal, label=label)
    ggplot (ggplotDF, aes(x=time, y=signal, colour=label)) + geom_line()
}

if (require(ggplot2)) {
    CSignals <- readEdfSignals (CHdr)
    plotEdfSignals (CSignals, labels=c('sine 8.5 Hz', 'sine 50 Hz'), from=.2, till=0.5)
}
```

For details about ggplot2 see e.g. the 'R Graphics cookbook'

Enjoy.

# 7. Acknowledgement
This package has used code from

- edf.R version 0.3 (27-11-2013) from Fabien Feschet    

- the work of Henelius Andreas as of July 2015,     
https://github.com/bwrc/edf

# 8. References

1. Specification of EDF        
http://www.edfplus.info/specs/edf.html

2. Specification of EDF+    
http://www.edfplus.info/specs/edfplus.html

3. Specification of BDF   
see 'Which file format does BioSemi use' at     
http://www.biosemi.com/faq/file_format.htm

4. Specification of BDF+       
http://www.teuniz.net/edfbrowser/bdfplus%20format%20description.html

Other useful EDF related sources can be found at:   
http://www.edfplus.info/downloads/
